Random number generation and applications on GPU’s in R?

### 2021-09-28 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
better simulation number in fisher and solved a concern
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2869 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Creating streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states (see the first column in the \proglang{R} output below): the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<streams on CPU>>=
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(Nstreams=4)
t(streams_on_CPU)
@
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<streams on GPU1, eval=TRUE>>=
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
@ 
<<streams on GPU2, eval=TRUE, echo=FALSE>>=
gpuR::colnames(streams_from_GPU) = c("current.g1.1", "current.g1.2", "current.g1.3", "current.g2.1", "current.g2.2", "current.g2.3",
"initial.g1.1", "initial.g1.2", "initial.g1.3", "initial.g2.1", "initial.g2.2", "initial.g2.3",
"substream.g1.1", "substream.g1.2", "substream.g1.3", "substream.g2.1", "substream.g2.2", "substream.g2.3")   
@ 
<<streams on GPU3,eval=TRUE>>=
t(streams_from_GPU)   
@
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
\fct{CreateStreamsGpu} in \pkg{gpuRandom} are similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

@@

We show the usage and advantage of \fct{gpuRandom::fisher.sim} by computing the p-values for two real data examples: one with a relatively big p-value and another with a very small p-value, and compare the runtime with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and an ordinary GPU, the other is equipped with an excellent GPU and an ordinary CPU. The device information is included in table \ref{tab:device}. The \proglang{R} outputs for testing on computer 2 is shown in the following. 
<<deviceinfo, echo=FALSE,cache=TRUE>>=
library(kableExtra)
a <- data.frame(
  var1 = c(" ", "Computer1", "Computer2"),
  var2 = c('Host', 'Intel Xenon W-2145 3.7Ghz', 	'VCPU Intel Xenon Skylake 2.5Ghz'),
  var3 = c('Device', 'AMD Radeon VII', 'VGPU Nvidia Tesla V100')
)

@@

knitr::kable(dt, col.names = NULL, caption = "Summary of comparions of Fisher's test simulation on different devices.\\label{tab:summary}") %>%
  kable_styling(full_width = FALSE, position = "center",latex_options = "HOLD_position") %>%
  kableExtra::group_rows(index = c("P-value" = 2, "Run-time" = 2)) %>%
  add_header_above(c(" " = 1, "Computer 1" = 2, "Computer 2" = 2, " " = 1))
@
% \begin{table}[H]
% \centering
% \begin{tabular}{ |c|c|c|c|c|c| } 
%  \hline
%  data &   B  & Intel CPU &  AMD GPU & NVIDIA GPU \\ \hline
%  Month & 1e6 & 0.403804 & 0.403507  & 0.403507     \\ \hline
%  Week &  1e7  & 0.0001251 & 0.0001274 & 0.0001274\\ \hline
% \end{tabular}
% \caption{ Summary of p-values obtained on different devices.\label{tab:summary}}
% \end{table}
% 
% 
% \begin{table}[H]
% \centering
% \begin{tabular}{ |c|c|c|c|c|c| } 
%  \hline
%  data &   B  & Intel CPU & sencha CPU & AMD GPU & NVIDIA GPU \\ \hline
%  Month & 1e6 & 9.74 & 14.58 & 2.28 & 0.31\\ \hline
%  Week &  1e7  & 60.19 & 90.38 & 10.82 & 1.90\\ \hline
% \end{tabular}
% \caption{ Run-time compare of Fisher's test simulation on different devices, all numbers are in seconds.\label{tab:compare time}}
% \end{table}

>>>>

\subsection{Creating streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states (see the first column in the \proglang{R} output below): the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<streams on CPU>>=
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(Nstreams=4)
t(streams_on_CPU)
@
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<streams on GPU1, eval=TRUE>>=
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
@ 
<<streams on GPU2, eval=TRUE, echo=FALSE>>=
gpuR::colnames(streams_from_GPU) = c("current.g1.1", "current.g1.2", "current.g1.3", "current.g2.1", "current.g2.2", "current.g2.3",
"initial.g1.1", "initial.g1.2", "initial.g1.3", "initial.g2.1", "initial.g2.2", "initial.g2.3")   
@ 
<<streams on GPU3,eval=TRUE>>=
t(streams_from_GPU)   
@
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
\fct{CreateStreamsGpu} in \pkg{gpuRandom} are similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

@@

We show the usage and advantage of \fct{gpuRandom::fisher.sim} by computing the p-values for two real data examples: one with a relatively big p-value and another with a very small p-value, and compare the runtime with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and an ordinary GPU, the other is equipped with an excellent GPU and an ordinary CPU. The \proglang{R} outputs for testing on computer 2 is shown in the following. 
<<deviceinfo, include=FALSE,cache=FALSE>>=
library(kableExtra)
a <- data.frame(
  var1 = c(" ", "Computer1", "Computer2"),
  var2 = c('Host', 'Intel Xenon W-2145 3.7Ghz', 	'VCPU Intel Xenon Skylake 2.5Ghz'),
  var3 = c('Device', 'AMD Radeon VII', 'VGPU Nvidia Tesla V100')
)

@@

knitr::kable(dt, col.names = NULL, caption = "Summary of comparions of Fisher's test simulation on different devices. Computer 1 is equipped with CPU Intel Xenon W-2145 3.7Ghz and AMD Radeon VII. Computer 2 is equipped with	VCPU Intel Xenon Skylake 2.5Ghz and VGPU Nvidia Tesla V100.\\label{tab:summary}") %>%
  kable_styling(full_width = FALSE, position = "center",latex_options = "HOLD_position") %>%
  kableExtra::group_rows(index = c("P-value" = 2, "Run-time" = 2)) %>%
  add_header_above(c(" " = 1, "Computer 1" = 2, "Computer 2" = 2, " " = 1))
@
% \begin{table}[H]
% \centering
% \begin{tabular}{ |c|c|c|c|c|c| } 
%  \hline
%  data &   B  & Intel CPU &  AMD GPU & NVIDIA GPU \\ \hline
%  Month & 1e6 & 0.403804 & 0.403507  & 0.403507     \\ \hline
%  Week &  1e7  & 0.0001251 & 0.0001274 & 0.0001274\\ \hline
% \end{tabular}
% \caption{ Summary of p-values obtained on different devices.\label{tab:summary}}
% \end{table}
% 
% 
% \begin{table}[H]
% \centering
% \begin{tabular}{ |c|c|c|c|c|c| } 
%  \hline
%  data &   B  & Intel CPU & sencha CPU & AMD GPU & NVIDIA GPU \\ \hline
%  Month & 1e6 & 9.74 & 14.58 & 2.28 & 0.31\\ \hline
%  Week &  1e7  & 60.19 & 90.38 & 10.82 & 1.90\\ \hline
% \end{tabular}
% \caption{ Run-time compare of Fisher's test simulation on different devices, all numbers are in seconds.\label{tab:compare time}}
% \end{table}
